Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-10-13 ms.tex
Brian revision to manuscripts, changes to vw-contour plot

The one-dimensional steady state flow of gas is usually characterized
by an inflow-outflow structure, with a critical radius known as the
``stagnation radius" $\rs$ where the radial velocity goes to zero.
Mass loss from stars interior to the stagnation radius is accreted,
while that outside $\rs$ is unbound in an outflow from the nucleus.
The stagnation radius, rather than the Bondi radius, is thus
ultimately responsible for setting the SMBH accretion rate, although
we find that $\rs$ typically lies within a factor of few of the
nominal Bondi radius.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion. 
The main goal of our analysis is to provide a framework for
interpreting the properties of quiescent galactic nuclei given the
relatively well understood physics of stellar winds and supernova
heating.

@@

Following \citet{Quataert:2004a} (see also \citealt{HolzerAxford:1970a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}), we calculate the gas density profile $\rho$ of the CNM for each galaxy within our sample by solving the equations of one-dimensional (spherically symmetric) time-dependent hydrodynamics,
\begin{align}
  &\drhodt+\frac{1}{r^2}\frac{\partial}{\partial r}\left(\rho r^2 v\right)=q \label{eq:drhodt}\\
  &\rho \left(\frac{\partial v}{\partial t} + v\frac{\partial v}{\partial r}\right) =-\dpdr-\rho \dphidr-q v \label{eq:dvdt}\\
  &\rho T\left(\frac{\partial s}{\partial t} + v\frac{\partial
      s}{\partial r}\right)=q\left[\ke+\kew-\gammaf \cs \right] 
  %+\nabla\cdot(\kappa \nabla T)
, 
\label{eq:dsdt}
\end{align}
where $v$, $T$, $p$ and $s$ are the radial velocity, temperature,
pressure and specific entropy of the gas, respectively.  Here $\partial
\Phi/\partial r = GM_{\rm enc}/r^{2}$ is the gravitational
acceleration, where $M_{\rm enc} = M_{\star}(r) + \Mbh$ is the
enclosed mass.  We assume an ideal gas equation of state with $p =
\rho kT/\mu m_p$ with $\mu = 1$  and $\gamma = 5/3$.

The source term in the continuity equation (equation~\ref{eq:drhodt}),
%Should equations be referenced as equation of equation
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 10^{10}$ years (see, e.g.,
\citealt{Ciotti+91}).  In Appendix \ref{app:eta} we estimate that
$\eta \approx 0.11(t_{\star}/t_{\rm h})^{-1.26}$ for a Salpeter IMF,
assuming that all stars are formed in a burst of star formation at a
common age of $t_{\star}$ (equation~[\ref{eq:eta}]).  We adopt
$\eta=1$ as fiducial, but note that changing $\eta$ is equivalent to
rescaling $\rho$ by the same factor.\footnote{This is {\it not} true
  if cooling is included, since the cooling rate is proportional to
  density {\it squared}.}

Source terms $\propto q$ also appear in the momentum and entropy
equations (equation~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) due to the fact
that the isotropic mass injection in the rest frame of the SMBH
represents a source of momentum and energy upon being incorporated
into the mean flow (see also below).  
%No net momentum is in the winds, but mass is being added which adds a
%term in the velocity equation...

% The second source term in
% equation (\ref{eq:dsdt}) accounts for heating transportation by
% electron conductivity, where the conductivity $\kappa = \kappa_0
% T^{5/2}$ is assumed to equal the classical ``Spitzer" value
% (\citealt{Spitzer62}); although even a weak magnetic field could in
% principle strongly suppress the conductivity perpendicular to the
% field, the radially-decreasing temperature profiles of interest are
% susceptible to the magneto-thermal instability (\citealt{Balbus01}),
% the non-linear evolution of which is believed to result in a
% radially-directed field structure (\citealt{Parrish&Stone07}).

Following \citet{ShcherbakovWong+:2014a}, the term in the entropy
equation $\propto \vw^2 = \sigma(r)^2+v_{w,0}^2$ accounts for external
heating sources.  Here $\sigma(r) = \sqrt{GM_{\rm enc}/r}$ is the
velocity dispersion of the stars, thus accounting for the fact that at
a minimum the gas experiences shock heating between the stellar winds
of velocities moving with respect to each other a characteristic
velocity $\sim \sigma$, i.e. even if the wind velocities themselves
are low, $\ll \sigma$, as characterize the majority of AGB mass loss.
The term $v_{w,0}^{2}$ parameterizes additional sources of heating,
such as main sequence winds, millisecond pulsars, supernovae, etc.  As
discussed in $\S\ref{sec:heating}$, the value of $v_{w,0}$ is
uncertain and will in general depend on the SMBH mass and the age of
the stellar population.

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability. This scheme is
described in \cite{Brandenburg:2003a}.  We solve the equations for
different choices of $v_{\rm w,0}$ spanning the physically expected
range.  Although we are most interested in the steady-state CNM
structure, the time-dependent equations are solved to avoid numerical
issues that arise near the critical sonic points, of which there are
up to two because of the supersonic inflow on small radial scales and
a possible supersonic outflow on large scales.

We do not include radiative cooling in the entropy equation because
across much of the parameter space radiative cooling is negligible
compared to other sources of cooling (e.g., advection).  Radiative
cooling cannot be neglected in some cases (especially if $\vwO$ is low
or $\eta$ is high), but the onset of a thermal instability when
radiative cooling dominates precludes obtaining a steady-state
solution anyways.  Our steady state solutions neglecting radiation can
nevertheless be used to ascertain over what parameter space cooling
instabilities are likely to develop ($\S\ref{sec:cooling}$).
%Dependence of the heating to cooling ratio has a far steeper
%dependence on vwO than on eta...

The accuracy of our numerical solutions is assessed by checking the
level of conservation of mass across the grid, as well as the integral
constraint on the energy.  Mass and energy are conserved to $\lesssim
40$ per cent for all solutions presented in this
paper. % {\bf BDM: more details...}

@@

This section provides analytic estimates for several quantities of
interest, such as the stagnation radius $\rs$ and the mass accretion
rate, the detailed derivation of which are given in Appendix
\ref{app:rs}.  All of our results are derived under the assumption
that thermal conductivity can be neglected.

When possible our results are expressed in terms of the wind heating
parameter $v_{\rm w,0}$; the power-law slope of the density profile
$n$ (also evaluated at $r = \rs$); and the dimensionless parameter \be
\chi \equiv \left(1 + \sigma^{2}/v_{\rm w,0}^{2}\right) \approx
\left(1+0.14 \Mbheight^{0.43}v_{500}^{-1}\right), \ee where $v_{500}
\equiv v_{w,0}/500$ km s$^{-1}$, $M_{\bullet,8} \equiv
M_{\bullet}/10^{8}M_{\odot}$, and the second equality makes use of the
$M_{\bullet}-\sigma$ correlation (\citealt{Gultekin+09}) \be
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\ee
Although the use of equation (\ref{eq:Msigma}) is questionable for low mass BHs (e.g., \citealt{Greene&Ho07}), our results are not particularly sensitive to this assumption.

First, by demanding continuity of the entropy derivative at the stagnation radius, the temperature at the stagnation radius is given by.
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p v_{w}^{2}}{2} \approx 6
\times 10^6\,v_{500}^2 \, \mu  \,\,{\rm K} 
\label{eq:Tanalytic}
\end{align}
%{\bf BDM: i picked $\mu = 0.62$ and stick to it.}

Next, by demanding hydrostatic equilibrium at the stagnation point, an expression for the stagnation radius can be derived,
\begin{align}
  \x=\frac{1}{\zeta^2 n}\left[
    \frac{\Mstar}{\Mbh}\left(A-\frac{n}{2}\right)+(B-\frac{n}{2})\right],
  \label{eq:stag_analytic}
\end{align}
where $\zeta \equiv v_{w,0}/\sigma$, $r_{\rm soi}
=GM_{\bullet}/\sigma^{2}$ is the radius of the SMBH sphere of
influence (soi), and A and B are constants that depend on the Nuker
parameter $\Gamma$ and the adiabatic index of the gas.

If the SMBH dominates the potential at the stagnation radius
($M_{\bullet} \gg M_{\star}$; $v_{w,0} \gg 200 M_{\bullet,8}^{0.2}$ km
s$^{-1}$) then the relationship between $\rs$ and $v_{w}$ is
simplified greatly:
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \simeq 6 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where we have assumed $n = 1$.  Note that this expression is similar to that
obtained by \citet{Volonteri+11} on heuristic grounds (their equation~[6]).

The SMBH accretion rate is simply given by the integrated mass loss rate interior to the stagnation radius (equation~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    1.6 \times 10^{-3}  M_{\bullet,8}^{1.9}
    v_{500}^{-4}  \eta \Msun \, \pyear& \text{core} \\
    3.4 \times 10^{-3} \Mbheight^{1.4}
    v_{500}^{-2}  \eta \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where the numerical estimates again assume that $M_{\bullet} \gg M_{\star}$ at the stagnation radius.  Note that equation (\ref{eq:mdot_analytic}) predicts that $\dot{M}$ should increase with BH mass at fixed $v_{500}$ but that $\dot{M}$ is extremely sensitive to the assumed heating rate $\propto v_{500}$.

Finally, the density at the stagnation radius $\rho(r_{\rm s})$ can be estimated by using equations (\ref{eq:stag_simple}) and (\ref{eq:mdot_analytic}) in conjunction with an alternative estimate of the mass accretion rate as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}(\rs) = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$, viz.
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{t_{\ff}(\rs)},
  \label{eq:mdot_gas}
\end{align}
from which one derives 
\begin{align}
  \rho(\rs)\simeq
  \begin{cases}
    3 \times 10^{-24} \Mbheight^{-0.14} v_{500}^{-1}  \eta \,
    \, {\rm g \, cm^{-3}}& \text{core}\\
    6 \times 10^{-24}  \Mbheight^{-0.56} v_{500}    \eta \,\, {\rm g \,cm^{-3}} & \text{cusp}
  \end{cases}
  \label{eq:rhors}
\end{align}

@@

Figure~\ref{fig:profiles} shows three representative profiles of the
density $\rho(r)$, temperature $T(r)$, and radial velocity $|v(r)|$.
As expected, the gas density increases towards the SMBH.  For some
profiles (e.g. NGC1172 and NGC4478) the density $\rho$ follows a power
law from the stagnation radius (squares in Fig.~\ref{fig:profiles})
inwards $\rho \propto r^{n}$ with $n \approx 1$. Note that this is
shallower than the $-3/2$ power law for Bondi accretion. However, the
power law behavior does not extend through all radii, and the gas
density profile may have breaks mirroring those in the stellar density
profile. For example, NGC4365 has a prominent break in the gas density
profile, near the stagnation radius. This coincides with the location
of the break in the stellar light profile ($\rb$). The temperature
profile is relatively flat at large radii, but increases as $\propto
1/r^{k}$ interior to the sphere of influence, where $k$ is smaller
than 1--the power expected for virialized gas.  The inwardly
directed velocity increases towards the hole approaching the expected
free-fall rate $v \propto v_{\rm ff}\propto r^{-1/2}$ interior to the
sonic point, and then increases outwards towards large radii.
% {\bf BDM: I made this up - in general we need
%   more explanation of what the plots are actually showing, with some
%   physical explanation.}
% AG -- things are not quite approaching Bondi on the inner grid. I
% don't have a good physical explanation yet...
% AG -- I am somewhat concerned about boundary effects in
% NGC3115. Perhaps the break in the gas density profile is due to
% boundary effects--I know that in this case the break is not
% associated with a break in the stellar density profile

@@

Though representing a much smaller quantity of mass, main sequence
winds have much higher velocities and hence contribute a
disproportionally large share of the kinetic energy of winds.
\citep{NaimanSoares-Furtado+:2013a} estimate that $v_{\rm w,0} \sim
100-150$ km s$^{-1}$ for main sequence winds assuming a relatively old
stellar population.  On the other hand, for a very young stellar
population of age $t_{\star} \lesssim 10^{7}$ yrs the effective wind
heating from massive stars can be much greater, $\gtrsim 10^{3}$ km
s$^{-1}$.  
%In Appendix C we present a toy model for how $v_{\rm w,0}$
%varies as a function of stellar age.
%AG-Appendix C is not written! As this draft will be circulated I
%don't want references to things which do not exist!

\subsubsection{Millisecond Pulsars} The spin down power of millisecond
pulsars (MSPs) could in principle be an important heating source.  If
the number of MSPs per unit stellar mass is $n_{\rm msp}$ and each
contributes an average spin-down luminosity $L_{\rm sd}$, then the
resulting heating per unit volume $e \approx L_{\rm sd}n_{\rm
  msp}\epsilon$ corresponds to an effective heating rate (equation~
[\ref{eq:vw_eff}])
  \begin{align}
    \vwO=\sqrt{\frac{\epsilon 2 t_h n_{\rm msp} L_{\rm sd}}{\eta}},
  \end{align}
where $\epsilon< 1$ is the thermalization efficiency of the wind.  The $\sim 30,000$ estimated MSPs in the Milky Way of stellar mass $6\times 10^{10}M_{\odot}$ (\citealt{Lorimer13}) corresponds to $n_{\rm msp}\simeq3 \times 10^{-40} $ MSPs g$^{-1}$.  If we assume an average spin-down luminosity $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic of Fermi pulsars, then we find that $\vwO\simeq 120\epsilon^{1/2} (\eta/0.1)^{-1/2}$ km s$^{-1}$.  Thus, assuming a high thermalization efficiency $\epsilon \sim 1$, MSP heating can be comparable to main sequence heating.  However, much lower thermalization efficiencies of $\epsilon \lesssim 0.1$ are inferred based on modeling the insterstellar mediums of globular clusters \citep{NaimanSoares-Furtado+:2013a}, while the binaries giving rise to through accretion MSPs may be disocciated by stellar interactions in the dense nuclear cluster.  
\subsubsection{Type Ia Supernovae}

Energy injected by Type Ia supernovae (SNe Ia) represents a potentially important source of heating, which unlike core collapse SNe is present even in an evolved stellar population.  Assuming that each Ia injects a thermal energy $E_{\rm Ia}$ into the instellar medium, and that the SNe rate (per stellar mass) is given by $R_{\rm Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm Ia}$ produces an effective wind heating parameter (equation~[\ref{eq:vw_eff}])
\begin{align}
    \vwO=\sqrt{\frac{2 t_h R_{\rm Ia} E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align}
The thermal energy injected by a Ia SN $E_{\rm Ia} \simeq \epsilon_{\rm Ia} 10^{51}$ ergs depends on the efficiency $\epsilon$ with which the blast wave energy is converted into bulk or turbulent motion instead of being lost to radiation.  \cite{Thornton+98} estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$, depending weakly on surrounding density, but \citet{Sharma+14} argues that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if the SNe occur in a hot dilute medium, as may characterize the CNM.

>>>>

The one-dimensional steady state flow of gas is usually characterized
by an inflow-outflow structure, with a critical radius known as the
``stagnation radius" $\rs$ where the radial velocity goes to zero.
Mass loss from stars interior to the stagnation radius is accreted,
while that outside $\rs$ is unbound in an outflow from the nucleus.
The stagnation radius, rather than the Bondi radius, is thus
ultimately responsible for setting the SMBH accretion rate, although
$\rs$ can in many cases lie within a factor of few of the
nominal Bondi radius.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion. 
The main goal of our analysis is to provide a framework for
interpreting the properties of quiescent galactic nuclei given the
comparatively well understood physics of stellar winds and supernova
heating.

@@

Following \citet{Quataert:2004a} (see also \citealt{HolzerAxford:1970a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}), we calculate the gas density profile $\rho$ of the CNM for each galaxy within our sample by solving the equations of one-dimensional (spherically symmetric) time-dependent hydrodynamics,
\begin{align}
  &\drhodt+\frac{1}{r^2}\frac{\partial}{\partial r}\left(\rho r^2 v\right)=q \label{eq:drhodt}\\
  &\rho \left(\frac{\partial v}{\partial t} + v\frac{\partial v}{\partial r}\right) =-\dpdr- \rho\frac{GM_{\bf enc}}{r^{2}} -q v \label{eq:dvdt}\\
  &\rho T\left(\frac{\partial s}{\partial t} + v\frac{\partial
      s}{\partial r}\right)=q\left[\ke+\kew-\gammaf \cs \right] +\nabla\cdot(\kappa \nabla T)
, 
\label{eq:dsdt}
\end{align}
where $v$, $T$, $p$ and $s$ are the radial velocity, temperature,
pressure and specific entropy of the gas, respectively, and $M_{\rm enc} = M_{\star}(r) + \Mbh$ is the
enclosed mass.  We assume an ideal gas equation of state with $p =
\rho kT/\mu m_p$ with $\mu = 0.62$  and $\gamma = 5/3$. {\bf BDM: take $\mu = 0.62$ here for fully ionized gas}

The source term in the continuity equation (equation~\ref{eq:drhodt}),
%Should equations be referenced as equation of equation
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 10^{10}$ years (see, e.g.,
\citealt{Ciotti+91}).  In Appendix \ref{app:eta} we estimate that
$\eta \approx 0.11(t_{\star}/t_{\rm h})^{-1.26}$ for a Salpeter IMF,
assuming that all stars are formed in a burst of star formation at a
common age of $t_{\star}$ (equation~[\ref{eq:eta}]).  We adopt
$\eta=0.1$ as fiducial {\bf BDM: take $\eta = 0.1$ instead of 1}, but note that changing $\eta$ is equivalent to
rescaling $\rho$ by the same factor.\footnote{This is {\it not} true
  if cooling is included, since the cooling rate is proportional to
  density {\it squared}.}

Source terms $\propto q$ also appear in the momentum and entropy
equations (equation~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) due to the fact
that the isotropic mass injection in the rest frame of the SMBH
represents a source of momentum and energy relative to the mean flow (see also below).  
%No net momentum is in the winds, but mass is being added which adds a
%term in the velocity equation...

The second source term in equation (\ref{eq:dsdt}) accounts for heating transportation by electron conductivity, where the conductivity $\kappa = \kappa_0 T^{5/2}$ is assumed to equal the classical ``Spitzer" value (\citealt{Spitzer62}); although even a weak magnetic field could in principle strongly suppress the conductivity perpendicular to the field, the radially-decreasing temperature profiles of interest are susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field structure (\citealt{Parrish&Stone07}).

Following \citet{ShcherbakovWong+:2014a}, the term in the entropy
equation $\propto \vw^2 = \sigma(r)^2+v_{w,0}^2$ accounts for external
heating sources.  Here $\sigma(r) = \sqrt{GM_{\rm enc}/r}$ is the
velocity dispersion of the stars, thus accounting for the fact that at
a minimum the gas experiences shock heating between the stellar winds
of velocities moving with respect to each other a characteristic
velocity $\sim \sigma$.  This heating is present if the wind velocities themselves
are low, $\ll \sigma$, as characterize the majority of AGB mass loss.
The term $v_{w,0}^{2}$ parameterizes additional sources of heating,
such as main sequence winds, millisecond pulsars, supernovae, etc.  As
discussed in $\S\ref{sec:heating}$, the value of $v_{w,0}$ is
uncertain and will in general depend on the SMBH mass and the age of
the stellar population.

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability (\citealt{Brandenburg:2003a}).  We solve the equations for
different choices of $v_{\rm w,0}$ spanning the physically expected
range.  Although we are most interested in the steady-state CNM
structure, the time-dependent equations are solved to avoid numerical
issues that arise near the critical sonic points, of which there are
up to two because of the supersonic inflow on small radial scales and
a possible supersonic outflow on large scales.

We do not include radiative cooling in the entropy equation because
across much of the parameter space radiative cooling is negligible
compared to other sources of cooling (e.g., advection).  Although radiative
cooling cannot be neglected in some cases (especially if $\vwO$ is low
or $\eta$ is high), our 1D scheme cannot account for the true multi-phase behavior of local thermal instabilities anyways (e.g.~\citealt{textit}).  Our steady state solutions neglecting radiation can
nevertheless be used to ascertain over what parameter space cooling
instabilities are likely to develop ($\S\ref{sec:cooling}$).
%Dependence of the heating to cooling ratio has a far steeper
%dependence on vwO than on eta...

The accuracy of our numerical solutions is assessed by checking the
level of conservation of mass across the grid, as well as the integral
constraint on the energy.  Mass and energy are conserved to $\lesssim
40$ per cent for all solutions presented in this
paper. 
{\bf BDM: more details...}

@@

This section summarizes analytic estimates of several physical quantities of interest, such as the stagnation radius $\rs$ and the mass accretion rate, the detailed derivation of which are given in Appendix \ref{app:rs}.  All of our results are derived under the assumption that thermal conductivity can be neglected.

When possible our results are expressed in terms of the wind heating
parameter $v_{\rm w,0}$; the power-law slope of the density profile
$n$ (also evaluated at $r = \rs$); and the dimensionless parameter \be
\chi \equiv v_{\rm w}^{2}/v_{\rm w,0}^{2} \approx \left(1 + \sigma^{2}/v_{\rm w,0}^{2}\right) \approx
\left(1+0.14 \Mbheight^{0.43}v_{500}^{-1}\right), \ee where $v_{500}
\equiv v_{w,0}/500$ km s$^{-1}$, $M_{\bullet,8} \equiv
M_{\bullet}/10^{8}M_{\odot}$, and the second equality makes use of the
$M_{\bullet}-\sigma$ correlation (\citealt{Gultekin+09}) \be
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\ee
Although the use of equation (\ref{eq:Msigma}) is questionable for low mass BHs (e.g., \citealt{Greene&Ho07}), our results are not particularly sensitive to this assumption.

First, by demanding continuity of the entropy derivative at the stagnation radius, the temperature at the stagnation radius is given by.
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p v_{w}^{2}}{2} \approx 3.7
\times 10^6\,\chi v_{500}^2 \,\,{\rm K} 
\label{eq:Tanalytic}
\end{align}
%{\bf BDM: i picked $\mu = 0.62$ and stick to it.}

Demanding hydrostatic equilibrium at the stagnation point provides an expression for the stagnation radius
\begin{align}
  \x=\frac{1}{\zeta^2 n}\left[
    \frac{\Mstar}{\Mbh}\left(A-\frac{n}{2}\right)+(B-\frac{n}{2})\right],
  \label{eq:stag_analytic}
\end{align}
where $\zeta \equiv v_{w,0}/\sigma$, $n$ is the power-law slope of the
density profile at $r = r_{\rm s}$,

\begin{align}
r_{\rm soi} = \frac{GM_{\bullet}}{\sigma^{2}} \approx 14M_{\bullet,8}^{0.6}\,\,{\rm pc}
\end{align}

is the radius of the SMBH sphere of
influence (soi), and A and B are constants that depend on the Nuker
parameter $\Gamma$ and the adiabatic index of the gas.

If the SMBH dominates the potential at the stagnation radius
($M_{\bullet} \gg M_{\star}$; $v_{w,0} \gg 200 M_{\bullet,8}^{0.2}$ km
s$^{-1}$) then the relationship between $\rs$ and $v_{w}$ is
simplified greatly:
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \simeq 6 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where in the numerical estimate we have assume $n = 1$.  This expression is similar to that obtained by \citet{Volonteri+11} on heuristic grounds (their equation~[6]).  On the other hand, if wind heating is much weaker ($v_{\rm w,0} \ll \sigma$), the stagnation radius moves to very large radii or may not exist at all.

Though important across some radii, the importance of conductive heating can be estimated by comparing its magnitude to that of wind heating, viz.~
\begin{eqnarray}
\left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm w}^{2}/2}\right|_{r_{\rm s}} &\sim& \frac{2t_{\rm h}\kappa_0 T^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}(r_{\rm s}) v_{\rm w}^{2}} \nonumber \\
&\sim& 0.5\left(\frac{\eta}{0.1}\right)^{-1}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-\beta} v_{500}^{3}\chi^{3},
\end{eqnarray}
where we have approximated $\nabla^{2} \sim 1/r_{\rm st}^{2}$ and in the second equality have made use of equations (\ref{eq:Tanalytic}), (\ref{eq:stag_simple}) and have approximated $4\pi r_{\rm s}^{3}\rho_{\star}(r_{\rm s}) \approx M_{\bullet}(r_{\rm s}/r_{\rm soi})^{\beta}$ with $\beta \approx 1-2$.  We thus see that conductive heating is important near the stagnation radius for high values of $v_{\rm w} \gtrsim 500$ km s$^{-1}$ and $\eta \gtrsim 0.1$, but that it is safely neglected to first order for lower heating rates.

The SMBH accretion rate is simply given by the integrated mass loss rate interior to the stagnation radius (equation~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    1.6 \times 10^{-3}  M_{\bullet,8}^{1.9}
    v_{500}^{-4}  \eta \Msun \, \pyear& \text{core} \\
    3.4 \times 10^{-3} \Mbheight^{1.4}
    v_{500}^{-2}  \eta \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where the numerical estimates again assume that $M_{\bullet} \gg M_{\star}$ at the stagnation radius.  Equation (\ref{eq:mdot_analytic}) predicts that $\dot{M}$ should increase with SMBH mass at fixed heating $v_{500}$, but decreases rapidly with increased heating.

The density of gas at the stagnation radius $\rho(r_{\rm s})$ can be estimated by using equations (\ref{eq:stag_simple}) and (\ref{eq:mdot_analytic}) in conjunction with an alternative estimate of the mass accretion rate as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}(\rs) = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$, viz.
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{t_{\ff}(\rs)},
  \label{eq:mdot_gas}
\end{align}
from which one derives 
\begin{align}
  \rho(\rs)\simeq
  \begin{cases}
    3 \times 10^{-24} \Mbheight^{-0.14} v_{500}^{-1}  \eta \,
    \, {\rm g \, cm^{-3}}& \text{core}\\
    6 \times 10^{-24}  \Mbheight^{-0.56} v_{500}    \eta \,\, {\rm g \,cm^{-3}} & \text{cusp}
  \end{cases}
  \label{eq:rhors}
\end{align}

@@

Figure~\ref{fig:profiles} shows three representative profiles of the density $\rho(r)$, temperature $T(r)$, and radial velocity $|v(r)|$.  As expected, the gas density increases towards the SMBH.  For some profiles (e.g. NGC1172 and NGC4478) the density $\rho$ follows a power law from the stagnation radius (squares in Fig.~\ref{fig:profiles}) inwards $\rho \propto r^{-n}$ with $n \approx 1$. Note that this is shallower than the $-3/2$ power law for Bondi accretion. However, the power law behavior does not extend through all radii, and the gas density profile may have breaks mirroring those in the stellar density profile. For example, NGC4365 has a prominent break in the gas density profile, near the stagnation radius. This coincides with the location
of the break in the stellar light profile ($\rb$). The temperature profile is relatively flat at large radii, but increases as $\propto
1/r^{k}$ interior to the sphere of influence, where $k$ is smaller than 1--the power expected for virialized gas.  The inwardly
directed velocity increases towards the hole approaching the expected free-fall rate $v \propto v_{\rm ff}\propto r^{-1/2}$ interior to the sonic point, and then increases outwards towards large radii.
% {\bf BDM: I made this up - in general we need
%   more explanation of what the plots are actually showing, with some
%   physical explanation.}
% AG -- things are not quite approaching Bondi on the inner grid. I
% don't have a good physical explanation yet...
% AG -- I am somewhat concerned about boundary effects in
% NGC3115. Perhaps the break in the gas density profile is due to
% boundary effects--I know that in this case the break is not
% associated with a break in the stellar density profile

@@

Main sequence winds, though contributing a much smaller quantity of mass, have much higher velocities and hence contribute a
disproportionally large share of the kinetic energy of winds.  \citet{NaimanSoares-Furtado+:2013a} estimate that $v_{\rm w,0} \sim 100-150$ km s$^{-1}$ for main sequence winds assuming a relatively old stellar population.  On the other hand, for a very young stellar population of age $t_{\star} \lesssim 10^{7}$ yrs the effective wind heating from massive stars can be much greater, $\gtrsim 10^{3}$ km s$^{-1}$.  In what follows below, we take an effective stellar wind heating rate
\be
v_{\rm w,0}^{\star} =  150\,{\rm km\,s^{-1}}
\ee
In Appendix \ref{app:windheat} we present a toy model for how $v_{\rm w,0}$ varies as a function of stellar age ({\bf BDM: Nick can create}).

\subsubsection{Millisecond Pulsars} The spin down power of millisecond
pulsars (MSPs) could in principle be an important heating source.  If
the number of MSPs per unit stellar mass is $n_{\rm msp}$ and each
contributes an average spin-down luminosity $\bar{L}_{\rm sd}$, then the
resulting heating per unit volume $e \approx L_{\rm sd}n_{\rm
  msp}\epsilon$ corresponds to an effective heating rate (eq.~
[\ref{eq:vw_eff}])

  \begin{align}
    v_{\rm w,0}^{\rm p} &=&\left({\frac{2\epsilon t_h n_{\rm msp} \bar{L}_{\rm sd}}{\eta}}\right)^{1/2} \nonumber \\
&\approx&  40\epsilon^{1/2}\left(\frac{\bar{L}_{\rm sd}}{10^{34}\,{\rm erg\,s^{-1}}}\right)^{1/2} \left(\frac{\eta}{0.1}\right)^{-1/2}\,{\rm km\,s^{-1}},
  \end{align}

where $\epsilon< 1$ is the thermalization efficiency of the wind.  The $\sim 30,000$ estimated MSPs in the Milky Way of stellar mass $6\times 10^{10}M_{\odot}$ (\citealt{Lorimer13}) corresponds to $n_{\rm msp}\simeq3 \times 10^{-40} $ MSPs g$^{-1}$, a number we assume in the numerical estimate.  Using the ATNF radio pulsar catalog (\citealt{Manchester+05}), we estimate the average spin-down luminosity of millisecond pulsars in the field (excluding those in globular clusters) to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg s$^{-1}$, in which case $v_{\rm w,0}$ is negligibly small even for $\eta = 0.1$ and $\epsilon = 1$.

If we assume a higher spin-down luminosity $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic of some Fermi pulsars, then we find that $v_{\rm w,0}^{\rm p} \simeq 120\epsilon^{1/2} (\eta/0.1)^{-1/2}$ km s$^{-1}$, in which case MSP heating can be comparable to main sequence wind heating for $\eta = 0.1$, $\epsilon = 1$.  However, much lower thermalization efficiencies of $\epsilon \lesssim 0.1$ are inferred based on modeling the insterstellar mediums of globular clusters \citep{NaimanSoares-Furtado+:2013a}.  Also, the binaries giving rise to MSPs may be disocciated by stellar interactions in the dense nuclear cluster, reducing their numbers as compared to the field population.    
\subsubsection{Type Ia Supernovae}
